---
title: 'Reproducible Research: Peer Assessment 1'
author: "Jake Gundrum"
date: "February 1, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Loading and Preprocessing the data
first load packages that will be used
```{r load packages, results = "hide", warning = FALSE, message = FALSE}
library(dplyr)
library(ggplot2)
library(lubridate)
```

read the raw data
```{r read data}
data_input <- read.csv("./activity.csv", stringsAsFactors = FALSE)
```
Check the structure to see what variables need to be formatted or changed
```{r check structure}
str(data_input)
```
the date will need to be coerced into date format
```{r coerce date}
data_input$date <- as.Date(data_input$date)
```

## What is the mean number of steps taken per day?
create histogram of steps taken per day
```{r histogram of steps per day, message = FALSE, warning = FALSE}
data_input %>% group_by(date) %>%
               summarize(steps = sum(steps, na.rm = TRUE)) %>% 
               ggplot(aes(steps)) +
                    geom_histogram() +
                    ggtitle("Steps Taken per day")
```

calculate the total, median, and mean number of steps per day
```{r calculate steps per day}
summaryNum <- data_input %>% group_by(date) %>%
                             summarize(steps = sum(steps, na.rm = TRUE)) %>% 
                             as.data.frame() %>% 
                             summarize(meanSteps = mean(steps, na.rm = TRUE),
                             medianSteps = median(steps, na.rm = TRUE))
summaryNum
```
## What is the daily average activity pattern?
Group data by average activity at interval
```{r Group by Interval of Day}
data_input %>% group_by(interval) %>% 
               summarize(steps = mean(steps, na.rm = TRUE)) %>%
               as.data.frame() %>% 
                 ggplot(aes(x = interval, y = steps)) +
                 geom_line(col = "steelblue") +
                 ggtitle("Average Activity Throughout the Day")
```

Which interval has the highest step average?
```{r interval with highest step average}
data_input %>% group_by(interval) %>% 
               summarize(steps = mean(steps, na.rm = TRUE)) %>%
               as.data.frame() %>% 
               arrange(-steps) %>% 
               head(1)
```

## Imputing Missing Values
imput missing values based on their interval average
```{r impute missing values}
int_avg <- data_input %>% group_by(interval) %>% 
               summarize(steps = mean(steps, na.rm = TRUE)) %>%
               as.data.frame()
data_imputed <- data_input
for (ii in 1:nrow(data_imputed)) {
        if(!is.na(data_imputed[ii,"steps"])){next}
        else{
            data_imputed[ii,"steps"] <- int_avg[data_imputed[ii,"interval"] == int_avg[,"interval"], "steps"]
        }  
  }
```
Create a histogram with the imputed values to see if there is any difference
create histogram of steps taken per day
```{r histogram of steps per day (imputed data), message = FALSE, warning = FALSE}
data_imputed %>% group_by(date) %>%
               summarize(steps = sum(steps, na.rm = TRUE)) %>% 
               ggplot(aes(steps)) +
                    geom_histogram() +
                    ggtitle("Steps Taken per day (imputed)")
```
calculate the total, median, and mean number of steps per day with imputed data
```{r calculate steps per day (imputed)}
summaryImp <- data_imputed %>% group_by(date) %>%
                               summarize(steps = sum(steps, na.rm = TRUE)) %>% 
                               as.data.frame() %>% 
                               summarize(meanSteps = mean(steps, na.rm = TRUE),
                               medianSteps = median(steps, na.rm = TRUE))
summaryImp
```
As expected, the imputed data shows higher mean and median, this is because the daily step count is raised as a result of adding in unknowns that previously counted as zero when aggregating daily steps

## Are there differences in activity patters between weekdays and weekends?
Create a variable for day of week and weekend then look at the differences using the imputed data
```{r create variable and chart weekday/weekend}
data_imputed %>% mutate(dayOfWeek = wday(date, label = TRUE),
                        weekend   = ifelse(dayOfWeek %in% c("Sun", "Sat"),"Weekend","Weekday")) %>% 
                 group_by(weekend, interval) %>% 
                 summarize(steps = mean(steps, na.rm = TRUE)) %>% 
                 as.data.frame() %>%
                 ggplot(aes(x = interval, y = steps)) +
                 geom_line() +
                 facet_grid(weekend~.) +
                 ggtitle("Movement Pattern by Day of Week")
```

